Constraint Programming Algorithms used in Gecode

### 2014-05-13 paper.tex
corrections Michael

\begin{document}
	\title{Constraint Programming - Inside Gecode}
	\author{Benedikt~Schmidt}
	\markboth{Advanced Seminar for Security in Information Technology, Summer Term 2014}%
	{Benedikt Schmidt: Constraint Programming - Inside \emph{Gecode}}	
	\maketitle

@@

	\begin{figure}
		\center
		\begin{tikzpicture}
			\draw (0, 0) rectangle (1, 1);
			\node at (0, -0.3) {($x_1$, $y_1$)};
			\draw[arrows=<->] (1.2, 0) -- (1.2, 1);
			\node at (1.5, 0.5) {$d_1$};
			\draw (1.5, 1.5) rectangle(2, 2);
			\node at (1.5, 1.2) {($x_2$, $y_2$)};
			\draw[arrows=<->] (2.2, 1.5) -- (2.2, 2);
			\node at (2.5, 1.75) {$d_2$};
		\end{tikzpicture}
		\caption{Two squares which should not overlap}
		\label{fig:squares}
	\end{figure}

@@

	In the following I will explain the necessary steps to describe and solve a problem with constraint programming: Variables, constraints, search trees, constraint propagation and search algorithms. As example of an implementation I will refer to \emph{Gecode} \cite{gecode}.

	\section{State of the Art}
	\subsection{Variable Domains}
	In constraint programming a variable is connected to a set of possible values, the so called domain. Depending on the domain, a variable is called an integer, Boolean or floating variable. Possible ranges for a variable $x$ of each domain are:
	\begin{itemize}
		\item Integer: finite or infinite sets like $x \in \mathbb{N}$, $x \in \{-5, 4, 10\}$, \dots
		\item Bool: $x \in \{\text{true}, \text{false}\}$
		\item Float: intervals or unions of intervals like $x \in [-4, 3]$, $x \in [-10, 8] \cup [10, 12]$, \dots
	\end{itemize}
	This distinction is fundamental as especially for floating variables a combinatorial approach to constraint programming is impossible.

	\subsection{Constraint Types}
	Which statements are possible to make and how comfortable it is to do so varies from the programming language in which the problem is implemented. For sure, there is also a difference in the interfaces of the libraries for constraint programming. Typical constraints are
	\begin{itemize}
		\item Equalities: $x = y + 2$, $x^2 + y^2 = 1$, \dots
		\item Inequalities: $x \ne y$, \dots
		\item Relations: $x < 2 y$, $e^x < 4$, \dots
	\end{itemize}

	Different constraint types can also be combined, for example inequalities or equalities through logic operators to Boolean expressions. Some of these constraints may acquire special attention during the search for solutions to improve the runtime of the search. For this reason a lot of work has been done for example on the alldifferent constraint \refeq{eq:alldifferent} \cite{allDifferent}, which would be such a combined constraint.
	\begin{equation}
		\forall x_i, x_j \in D, i \ne j: x_i \ne x_j, 
		\label{eq:alldifferent}
	\end{equation}

@@

	In constraint programming, search trees, like the one in \reffig{fig:searchTree}, describe the partition of the search space. Their nodes have names, depending on their position in the tree. The search is started from the root node. At this point, all variable domains are still complete. The root node is also the first one with a branch, where the search space is split up into disjoint sets. This approach is similar to the divide-and-conquer technique, as it reduces a complex problem into two or more parts with less complexity. The splitting, which adds additional constraints on the variables, is then recursively done with different heuristics to partition the search space into smaller parts. If one of the parts is small enough to decide if it is feasible the node can be called a either a dead end or a solution. To end up as a dead end, there must be no feasible solutions left in the search space at this node. To turn the node into a solution it is necessary that all members of the reduced search space are valid considering the constraints. This can be the case if either only one choice is left for every variable, and the constraints are valid for this choice, or if a whole interval of a floating variable is valid for the constraints.

	\subsection{Constraint Propagation}
	Constraint propagation is a key concept in constraint programming as it reduces the size of the search space significantly. The term propagation refers in this area to the reduction of choices for variables which are infeasible, considering other constraints and affected variable domains. As the size of the variable domain is related to the possibilities which have to be checked by a search through additional subtrees, the constraint propagation is able to prune complete infeasible sub trees. Therefore, the whole search process can be accelerated through constraint propagation.

@@

	To illustrate this effect consider a constraint problem based on two variables $x \in \{1, 5, 6\}$ and $y \in \{1, 2\}$ and one constraint based on them.
	\begin{equation}
		c_1:\ x \le y
	\end{equation}
	Without constraint propagation and branching first on $x$, a decision, which may have been made by a certain branching strategy, the search trees turns out to be the one in \reffig{fig:bigTree}. By inspection one sees that the right parts of this tree can be already pruned at the root by propagation of constraint $c_1$. Therefore, it is not necessary to create and evaluate these sub trees. With constraint propagation fewer branches have to be made, like it can be seen in \reffig{fig:smallTree}.

@@

	Constraints in \emph{Gecode} are evaluated ahead of a further growth of a subtree, as it may be possible to prune some of the sub trees. The problem here is to decide in which order the constraints should be considered, and it may even be necessary to evaluate some constraints more often. A short example will illustrate this \cite[p.~575]{handbookCP}: Let there be two constraints 
	\begin{equation}
		c_1: y - x^2 = 0
	\end{equation}
	\begin{equation}
		c_2: y - x - 1 = 0
		\label{eq:linear}
	\end{equation}
	and let the variable domains be $x, y \in [-4, 4]$. Starting with the evaluation of constraint $c_1$ follows a chain of reductions of the variable domains, based only on constraint propagation:
	\begin{equation}
	\begin{split}
		c_1:\ &y \in [-4, 4] \rightarrow y \in [0, 4] \\
		c_1:\ &x \in [-4, 4] \rightarrow x \in [-2, 2] \\
		c_2:\ &y \in [0, 4] \rightarrow y \in [0, 3] \\
		c_2:\ &x \in [-2, 2] \rightarrow x \in [-1, 2] \\
		c_1:\ &x \in [-1, 2] \rightarrow x \in [-1, 1.73 \dots] \\
		&\dots
	\end{split}
	\end{equation}

	To implement this repeated evaluation of constraints in \emph{Gecode} all constraints subscribe to the domains of the variables which are part of them. With this method, the constraints get notified whenever a variable domain is reduced and they can signal the system that they need to be evaluated once again.

@@

		\subsection{Branching Strategies}
	If at one node it is not possible to empty one variable domain or to reduce all variable domains to the size of one it is necessary to make a decision: How should the next subtree be constructed? The question is answered by the selected branching strategy. No matter which one is selected, they all have in common that they create additional, temporary constraints, which are only applied to the specific sub trees. This process is called \emph{posting} a constraint on a branch.

	The branching strategy depends on the variable domain: For integer variables there are three common strategies \cite[p.~87]{handbookCP} which differ in how the domain is distributed to the sub trees.
	\begin{itemize}
		\item Enumeration: The domain for $x \in D = {x_1, x_2, \dots}$ is split up into $|D|$-branches where for every branch $i$ one additional constraint $c:\ x = x_i$ is posted.
		\item Binary choice points: One value $x_i$ from the domain $D$ of the variable $x$ is selected and the domain is split up into two sub trees: On one the constraint $x = x_i$ and on the other one the constraint $x \ne x_i$ is posted. This results into something similar to a binary tree at this point.
		\item Domain splitting: The domain $D$ for the variable $x$ is split up into two disjoint intervals or sets, for example $x < 2$ and $x \ge 2$.
	\end{itemize}

@@

	\subsubsection{Depth-First Search}
	The depth-first search was historical the first one which was developed, therefore it is a good point to start explaining search algorithms. The idea behind this search algorithm is that every possible combination of variables is examined. In its basic form it can not be applied to problems with infinite domains; for problems with such variable domains it is necessary to use for instance branch-and-reduce. Sometimes the depth-first search is also referred to as chronological backtrack search.

	The starting point for the depth-first search is the root node, from where on the algorithm is called recursively. At every node, mainly two steps are executed: First, a constraint propagation to reduce the variable domains and second a branching, which divides the search space into two disjoint sub spaces, on which the same algorithm is applied again. The search stops in a leaf of the search tree if either no branching is possible and therefore a valid solution is found, or one or more variable domains are emptied, which indicates that in this leave no possible solution exists.

@@

	By stopping the search after the first feasible solution is found the algorithm can often be accelerated. Although this slightly modified version is not complete anymore, in practice it is in some areas sufficient to know just one valid solution.

	\subsubsection{Depth-First Search with Optimization}
	If in the constraint programming an objective is defined which should be either minimized or maximized, a modified version of the depth-first search can be applied. The only difference is that it evaluates the objective function for every found solution concerning the constraints and stores during the search only the best one. To accelerate the search it is possible to add an additional constraint which defines that the new objective must better than the best so far. Through this enhancement even more sub trees are pruned and the search is faster.

	This algorithm is also a complete one. Hence, it will always find the best solution if one exists. This algorithm does not inherit the bad memory characteristics from the depth-first search as only one solution is stored, but the search is still exhaustive and can be very slow, depending on the size of the variable domains.

@@

			\draw[very thick] (a) -- (b) -- (c) -- (d) -- (e) -- (a);
			\draw (a) -- (b) -- (c) -- (e) -- (a);
			\draw (f) -- (g) -- (c) -- (b) -- (f);
			\draw[fill=blue, fill opacity=0.2] (c) -- (d) -- (e) -- (c);
			\draw[fill=green, fill opacity=0.2] (a) -- (f) -- (g) -- (e) -- (a);
			\draw[fill=brown, fill opacity=0.2] (f) -- (g) -- (c) -- (b) -- (f);
			\node at (2.75, 3.75) {$P_0$};
			\node at (1.25, 1.25) {$P_1$};
			\node at (3.75, 1.75) {$P_2$};
		\end{tikzpicture}	
		\caption{Splitting of a search space into three sub problems and the possible candidates for solutions}
		\label{fig:branchAndBound}
	\end{figure}

	The basic concept for finding solutions in branch-and-bound is an enumeration of all possible candidates, therefore it is necessary to separate the search space in a useful way, which is the task for the branch-and-reduce step. Actually this step is very similar to the concept of branching, it also depends on a certain heuristic to divide the search space into two or even more disjoint parts. This heuristic can either be selected by the user or be a general one, like splitting the biggest variable domain into two equal sized parts. If the basic search space contains a big amount of possible candidate solutions one reduction may not be enough and therefore this step is applied recursively to the newly create sub problems. This process can be stopped if the number of solutions is reduced to an amount, where it is possible to enumerate all of them and check if they are optimal. A possible division into sub problems can be seen in \reffig{fig:branchAndBound}, where for legibility branch-and-reduce was applied only twice: The first execution split the problem into $P_0$ and $P'_0$, which was then split again into $P_1$ and $P_2$.

	The next step after branch-and-reduce is the bounding, which is able to prune whole sub problems. In the following part I will consider the minimization of an objective function without a loss of generality:
	\begin{equation}
		\max_x \{f(x)\} = -\min_x \{-f(x)\}
	\end{equation}

@@

	The actual pruning of sub problems is based on the relaxed optimum, which is the optimum of the problem without consideration that only discrete values are allowed. If this relaxed optimum for a certain sub problem is greater than the current upper bound it is possible to prune this complete sub problem, as it can not contain a better solution.

	Consider for this \reffig{fig:branchAndBound}, where already a certain optimum $f$ was found in sub problem $P_1$. If now the relaxed optimum $f_{P0}$ in $P_0$ and $f_{P2}$ in $P_2$ is worse than this optimum
	\begin{equation}
		f_{P0} > f
	\end{equation}
	\begin{equation}
		f_{P2} > f
	\end{equation}
	the whole sub problems $P_{0}$ and $P_{2}$ can be pruned and it is not necessary to enumerate the solution candidates in them (or to split them further up).

@@

	\subsubsection{Limited Discrepancy Search}
	The limited discrepancy search falls into the category of best-first searches and was first mentioned by Harvey and Ginsberg \cite{limitedDiscrepancy}. It depends on branching heuristics for the tree, which are determined through branching strategies. The basic concept is that during a chronological backtrack search a wrong turn close to the root is first of all very expensive. Second, heuristics work worse the farther away from the solution they are applied, therefore decisions close to the root node are likely to fail. Consequently, in a depth-first search the worst decisions are the most expensive ones. The limited discrepancy search solves exactly this problem as it suggests to change rather the decisions made close to the root than the ones the search made later. The improvement compared to a chronological backtrack was shown by experimental results and theoretically proven in \cite{limitedDiscrepancy}.

@@

	\subsubsection{Restart Strategies}
	Branching always means to apply certain heuristics and these heuristics may fail or produce bad results for the first execution of the search algorithm. If it turns out that the search tree is build in a disadvantageous way a restart strategy may be appropriate to improve the search. For such a restart the heuristic is improved, based on the things the search has learned to be a not so good solution during the first or first few attempts to build a search tree.
	For practical application several variants of restart strategies were proposed \cite[p.~113]{handbookCP}

	\section{Conclusion}
	\emph{Gecode}, as representative for an implementation of constraint programming, provides the necessary tools to describe and solve a wide range of combinatorial or optimization problems. The interface allows an adaption to the area of application through modifications of algorithms, restarts or the implementation of a custom search algorithm, based on the already provided ones. Therefore \emph{Gecode} can be used in various areas, starting from scheduling, over portfolio optimization up to technical problems like placement.

	\begin{thebibliography}{1}
		\bibitem{handbookCP}
		F.~Rossi, P.~van~Beek and T.~Walsh, \emph{Handbook of Constraint Programming}, Amsterdam, The Netherlands: Elsevier, 2006, ISBN: 978-0-444-52726-4
		\bibitem{allDifferent}
		W.~van~Hoeve, \emph{The Alldifferent Constraint: A Survey}, http://www.andrew.cmu.edu/user/vanhoeve/papers/alldiff.pdf
		\bibitem{trendsInCP}
		F.~Benhamou, N.~Jussien, B.~A.~O'Sullivan, \emph{Trends in Constraint Programming}, Wiley-ISTE, 2007, ISBN: 978-1-905209-97-2
		\bibitem{gecode}
		http://www.gecode.org/
		\bibitem{programmingGecode}
		C.~Schulte, G.~Tack and M.~Z.~Lagerkvist, \emph{Modeling and Programming with Gecode}, http://www.gecode.org/doc-latest/MPG.pdf
		\bibitem{limitedDiscrepancy}
		W.~D.~Harvey, M.~L.~Ginsberg, \emph{Limited Discrepancy Search}, Morgan Kaufmann, 1995
		\bibitem{linearProgramming}
		L.~Cooper and D.~Steinberg, \emph{Methods and Applications of Linear Programming}, W.~B.~Saunder~Company, 1974, ISBN: 0-7216-2694-7
	\end{thebibliography}
\end{document}

>>>>

\begin{document}
	\title{Constraint Programming Algorithms used in Gecode}
	\author{Benedikt~Schmidt}
	\markboth{Advanced Seminar for Security in Information Technology, Summer Term 2014}%
	{Benedikt Schmidt: Constraint Programming - Inside \emph{Gecode}}	
	\maketitle

@@

	\begin{figure}[b]
		\center
		\begin{tikzpicture}
			\draw (0, 0) rectangle (1, 1);
			\node at (0, -0.3) {($x_1$, $y_1$)};
			\draw[arrows=<->] (1.2, 0) -- (1.2, 1);
			\node at (1.5, 0.5) {$d_1$};
			\draw (1.5, 1.5) rectangle(2, 2);
			\node at (1.5, 1.2) {($x_2$, $y_2$)};
			\draw[arrows=<->] (2.2, 1.5) -- (2.2, 2);
			\node at (2.5, 1.75) {$d_2$};
		\end{tikzpicture}
		\caption{Two squares which should not overlap}
		\label{fig:squares}
	\end{figure}

@@

	In the following I will explain the necessary steps to describe and solve a problem with constraint programming: variables, constraints, search trees, constraint propagation and search algorithms. As example of an implementation I will refer to \emph{Gecode} \cite{gecode}.

	\section{State of the Art}
	\subsection{Variable Domains}
	In constraint programming a variable is connected to a set of possible values, the so called domain. Depending on the domain, a variable is called an integer, Boolean or floating variable. Possible ranges for a variable $x$ of each domain are:
	\begin{itemize}
		\item Integer: finite or infinite, e.g. $x \in \mathbb{N}$, $x \in \{-5, 4, 10\}$
		\item Bool: $x \in \{\text{true}, \text{false}\}$
		\item Float: intervals or unions of intervals, e.g. $x \in [-4, 3]$, $x \in [-10, 8] \cup [10, 12]$
	\end{itemize}
	This distinction is fundamental as especially for floating variables a combinatorial approach to constraint programming is impossible.

	\subsection{Constraint Types}
	The choice of constraint types and how comfortable it is to specify them varies from the programming language in which the problem is implemented. For sure, there is also a difference in the interfaces of the libraries for constraint programming. Typical constraints are
	\begin{itemize}
		\item Equalities: $x = y + 2$, $x^2 + y^2 = 1$, \dots
		\item Inequalities: $x \ne y$, \dots
		\item Relations: $x < 2 y$, $e^x < 4$, \dots
	\end{itemize}

	Different constraint types can also be combined, for example inequalities or equalities through logic operators to Boolean expressions. Some of these constraints may acquire special attention during the search for solutions to improve the runtime of the search. For this reason a lot of work has been done for example on the alldifferent constraint \cite{allDifferent}
	\begin{equation}
		\forall x_i, x_j \in D, i \ne j: x_i \ne x_j, 
		\label{eq:alldifferent}
	\end{equation}
	which is an example for a combined constraint.

@@

	In constraint programming, search trees, like the one in \reffig{fig:searchTree}, describe the partition of the search space. Their nodes have names, depending on their position in the tree. The search is started from the root node. At this point, all variable domains are still complete. The root node is also the first one with a branch, where the search space is split up into disjoint sets. This approach is based on the same idea like the divide-and-conquer technique \cite[p.~175]{artOfComputerProgramming}, as it reduces a complex problem into two or more parts with less complexity. The splitting, which adds additional constraints on the variables, is then recursively done with different heuristics to partition the search space into smaller parts. If one of the parts is small enough to decide if it is feasible the node can be called a either a dead end or a solution. To end up as a dead end, there must be no feasible solutions left in the search space at this node. To turn the node into a solution it is necessary that all members of the reduced search space are valid considering the constraints. This can be the case if either only one choice is left for every variable, and the constraints are valid for this choice, or if a whole interval of a floating variable is valid for the constraints.

	\subsection{Constraint Propagation}
	Constraint propagation is a key concept in constraint programming as it reduces the size of the search space significantly. The term propagation refers in this area to the reduction of choices for variables which are infeasible, considering other constraints and affected variable domains. As the size of the variable domain is related to the possibilities which have to be checked by a search through additional subtrees, the constraint propagation is able to prune complete infeasible subtrees. Therefore, the whole search process can be accelerated through constraint propagation.

@@

	To illustrate this effect consider a problem based on two variables $x \in \{1, 5, 6\}$ and $y \in \{1, 2\}$ and the constraint
	\begin{equation}
		c_1:\ x \le y
	\end{equation}
	Without constraint propagation and branching first on $x$, a decision, which may have been made by a certain branching strategy, the search trees turns out to be the one in \reffig{fig:bigTree}. By inspection one sees that the right parts of this tree can be already pruned at the root by propagation of constraint $c_1$. Therefore, it is not necessary to create and evaluate these subtrees. With constraint propagation fewer branches have to be made, like it can be seen in \reffig{fig:smallTree}.

@@

	Constraints in \emph{Gecode} are evaluated ahead of a further growth of a subtree, as it may be possible to prune some of the subtrees. The problem here is to decide in which order the constraints should be considered, and it may even be necessary to evaluate some constraints more often. A short example will illustrate this \cite[p.~575]{handbookCP}: Let there be two constraints 
	\begin{equation}
		c_1: y - x^2 = 0
	\end{equation}
	\begin{equation}
		c_2: y - x - 1 = 0
		\label{eq:linear}
	\end{equation}
	and let the variable domains be $x, y \in [-4, 4]$. Starting with the evaluation of constraint $c_1$ follows a chain of reductions of the variable domains, based only on constraint propagation:
	\begin{equation}
	\begin{split}
		c_1:\ &y \in [-4, 4] \rightarrow y \in [0, 4] \\
		c_1:\ &x \in [-4, 4] \rightarrow x \in [-2, 2] \\
		c_2:\ &y \in [0, 4] \rightarrow y \in [0, 3] \\
		c_2:\ &x \in [-2, 2] \rightarrow x \in [-1, 2] \\
		c_1:\ &x \in [-1, 2] \rightarrow x \in [-1, 1.73 \dots] \\
		&\dots
	\end{split}
	\end{equation}

	To implement this repeated evaluation of constraints in \emph{Gecode} all constraints subscribe to the domains of the variables which are part of them. A more detailed explanation of this observer pattern can be found in \cite{designPatterns}. With this method, the constraints get notified whenever a variable domain is reduced and they can signal the system that they need to be evaluated once again.

@@

		\subsection{Branching Strategies}
	If at one node it is not possible to empty one variable domain or to reduce all variable domains to the size of one it is necessary to make a decision: How should the next subtree be constructed? The question is answered by the selected branching strategy. No matter which one is selected, they all have in common that they create additional, temporary constraints, which are only applied to the specific subtrees. This process is called \emph{posting} a constraint on a branch.

	The branching strategy depends on the variable domain: For integer variables there are three common strategies \cite[p.~87]{handbookCP} which differ in how the domain is distributed to the subtrees.
	\begin{itemize}
		\item Enumeration: The domain for $x \in D = {x_1, x_2, \dots}$ is split up into $|D|$-branches where for every branch $i$ one additional constraint $c:\ x = x_i$ is posted.
		\item Binary choice points: One value $x_i$ from the domain $D$ of the variable $x$ is selected and the domain is split up into two subtrees: On one the constraint $x = x_i$ and on the other one the constraint $x \ne x_i$ is posted. This results into something similar to a binary tree at this point.
		\item Domain splitting: The domain $D$ for the variable $x$ is split up into two disjoint intervals or sets, for example $x < 2$ and $x \ge 2$.
	\end{itemize}

@@

	\subsubsection{Depth-First Search}
	The idea behind this search algorithm is that every possible combination of variables is examined. In its basic form it can not be applied to problems with infinite domains; for problems with such variable domains it is necessary to use for instance branch-and-bound. Sometimes the depth-first search is also referred to as chronological backtrack search.

	The starting point for the depth-first search is the root node, from where on the algorithm is called recursively. At every node, mainly two steps are executed: First, a constraint propagation to reduce the variable domains and second a branching, which divides the search space into two disjoint subspaces, on which the same algorithm is applied again. The search stops in a leaf of the search tree if either no branching is possible and therefore a valid solution is found, or one or more variable domains are emptied, which indicates that in this leave no possible solution exists.

@@

	By stopping the search after the first feasible solution is found the runtime of the search can be improved. Although this slightly modified version is not complete anymore, in practice it is in some areas sufficient to know just one valid solution.

	\subsubsection{Depth-First Search with Optimization}
	If in the constraint programming an objective is defined which should be either minimized or maximized, a modified version of the depth-first search can be applied. The only difference is that it evaluates the objective function for every found solution concerning the constraints and stores during the search only the best one. To accelerate the search it is possible to add an additional constraint which requires that the new value for the objective must be better than the best achieved so far. Through this enhancement even more subtrees are pruned and the search is faster.

	This algorithm is also complete. Hence, it will always find the best solution, if one exists. This algorithm does not inherit the bad memory characteristics from the depth-first search as only one solution is stored, but the search is still exhaustive and can be very slow, depending on the size of the variable domains.

@@

			\draw[very thick] (a) -- (b) -- (c) -- (d) -- (e) -- (a);
			\draw (a) -- (b) -- (c) -- (e) -- (a);
			\draw (f) -- (g) -- (c) -- (b) -- (f);
			\draw[fill=blue, fill opacity=0.2] (c) -- (d) -- (e) -- (c);
			\draw[fill=green, fill opacity=0.2] (a) -- (f) -- (g) -- (e) -- (a);
			\draw[fill=brown, fill opacity=0.2] (f) -- (g) -- (c) -- (b) -- (f);
			\node at (2.75, 3.75) {$P_0$};
			\node at (1.25, 1.25) {$P_1$};
			\node at (3.75, 1.75) {$P_2$};
		\end{tikzpicture}	
		\caption{Splitting of a search space into three subproblems and the possible candidates for solutions}
		\label{fig:branchAndBound}
	\end{figure}

	The basic concept for finding solutions in branch-and-bound is an enumeration of all possible candidates. Therefore, it is necessary to separate the search space in an appropriate way, which is the task for the branch-and-reduce step. This step is very similar to the concept of branching, it also depends on a certain heuristic to divide the search space into two or even more disjoint parts. This heuristic can either be selected by the user or be a general one, like splitting the biggest variable domain into two equal sized parts. If the basic search space contains a big amount of possible candidate solutions one reduction may not be enough and therefore this step is applied recursively to the newly create subproblems. This process can be stopped if the number of solutions is reduced to an amount, where it is possible to enumerate all of them and check if they are optimal. A possible division into subproblems can be seen in \reffig{fig:branchAndBound}, where for legibility branch-and-reduce was applied only twice: The first execution split the problem into $P_0$ and $P'_0$, which was then split again into $P_1$ and $P_2$.

	The next step after branch-and-reduce is the bounding, which is able to prune whole subproblems. In the following part I will consider the minimization of an objective function without a loss of generality:
	\begin{equation}
		\max_x \{f(x)\} = -\min_x \{-f(x)\}
	\end{equation}

@@

	The actual pruning of subproblems is based on the relaxed optimum, which is the optimum of the problem without consideration that only discrete values are allowed. If this relaxed optimum for a certain subproblem is greater than the current upper bound it is possible to prune this complete subproblem, as it can not contain a better solution.

	Consider for this \reffig{fig:branchAndBound}, where already a certain optimum $f$ was found in subproblem $P_1$. If now the relaxed optimum $f_{P0}$ in $P_0$ and $f_{P2}$ in $P_2$ is worse than this optimum
	\begin{equation}
		f_{P0} > f
	\end{equation}
	\begin{equation}
		f_{P2} > f
	\end{equation}
	the whole subproblems $P_{0}$ and $P_{2}$ can be pruned and it is not necessary to enumerate the solution candidates in them (or to split them further up).

@@

	\subsubsection{Limited Discrepancy Search}
	The limited discrepancy search falls into the category of best-first searches and was first mentioned by Harvey and Ginsberg \cite{limitedDiscrepancy}. It depends on branching heuristics for the tree, which are determined through branching strategies. The basic concept is that during a chronological backtrack search a wrong turn close to the root is first of all very expensive. Second, heuristics work worse the farther away from the solution they are applied. Therefore, decisions close to the root node are likely to fail. Consequently, in a depth-first search the worst decisions are the most expensive ones. The limited discrepancy search solves exactly this problem as it suggests to change rather the decisions made close to the root than the ones the search made later. The improvement compared to a chronological backtrack was shown by experimental results and theoretically proven in \cite{limitedDiscrepancy}.

@@

	\subsubsection{Restart Strategies}
	Branching always means to apply certain heuristics and these heuristics may fail or produce bad results for the first execution of the search algorithm. If it turns out that the search tree is build in a disadvantageous way a restart strategy may be appropriate to improve the search. For such a restart the heuristic is improved, based on the things the search has learned to be a not so good solution during the first or first few attempts to build a search tree.
	For practical application several variants of restart strategies were proposed \cite[p.~113]{handbookCP}.

	\section{Conclusion}
	\emph{Gecode}, as representative for an implementation of constraint programming, provides the necessary tools to describe and solve a wide range of combinatorial or optimization problems. The interface allows an adaption to the area of application through modifications of algorithms, restarts or the implementation of a custom search algorithm, based on the already provided ones. Therefore, \emph{Gecode} can be used in various areas, starting from scheduling, over portfolio optimization up to technical problems like placement.

	\begin{thebibliography}{1}
		\bibitem{handbookCP}
		F.~Rossi, P.~van~Beek and T.~Walsh, \emph{Handbook of Constraint Programming}, Amsterdam, The Netherlands: Elsevier, 2006, ISBN-13: 978-0-444-52726-4
		\bibitem{allDifferent}
		W.~van~Hoeve, \emph{The Alldifferent Constraint: A Survey}, http://www.andrew.cmu.edu/user/vanhoeve/papers/alldiff.pdf
		\bibitem{trendsInCP}
		F.~Benhamou, N.~Jussien, B.~A.~O'Sullivan, \emph{Trends in Constraint Programming}, Wiley-ISTE, 2007, ISBN-13: 978-1-905209-97-2
		\bibitem{gecode}
		http://www.gecode.org/
		\bibitem{programmingGecode}
		C.~Schulte, G.~Tack and M.~Z.~Lagerkvist, \emph{Modeling and Programming with Gecode}, http://www.gecode.org/doc-latest/MPG.pdf
		\bibitem{limitedDiscrepancy}
		W.~D.~Harvey, M.~L.~Ginsberg, \emph{Limited Discrepancy Search}, Morgan Kaufmann, 1995
		\bibitem{linearProgramming}
		L.~Cooper and D.~Steinberg, \emph{Methods and Applications of Linear Programming}, W.~B.~Saunder~Company, 1974, ISBN-10: 0-7216-2694-7
		\bibitem{artOfComputerProgramming}
		D.~E.~Knuth, \emph{The Art of Computer Programming: Volume 3, Sorting and Searching}, second edition, Addison-Wesley, 1998, ISBN-10: 0201896850, ISBN-13: 9780201896855
		\bibitem{designPatterns}
		E.~Gamma, R.~Helm, R.~Johnson, J.~Vlissides, \emph{Design Patterns. Elements of Reusable Object-Oriented Software}, first edition, Prentice Hall, 1994, ISBN-10: 0201633612
	\end{thebibliography}
\end{document}
